Sketch Recognition using Enhanced Particle Swarm Optimization

### 2010-12-10 trunk/Swarm/swarmModified.tex

\section{Enhanced Particle Swarm}
\label{sec:EnhancedSwarmAlgorithm}
In the above basic PSO, the velocity update is based on three terms an inertia
term, local cognitive ($l_{best}$) term and global cognitive ($g_{best}$)term.
The particle local cognitive term is associated with the particle own history
without consideration of other swarm particles locations and histories which
leads to a lake of diversity and sub optimum solutions. Therefore, we are trying
to add more social influence in the particle movement by using the weighted
difference of the position vectors of any other two distinct particles randomly
chosen from the swarm to modify the local cognitive term. The modified term adds
social behavior of other particle and keeps diversity while searching for the
global optimum. The new social cognitive term of the particle is shifted to the
new location only if it yields a fitness value better than obtained using the
standard PSO\cite{stockPaper}.

@@

\section{System Discribtion  }\label{Sysdisc}
%\label{sec:application}
%The modified swarm algorithm is used to improve the performance of the sketch recognition system in \cite{mypaper}.

The sketch recognitions system is divided into three main steps 1) preprocessing, 2) segmentation and 3) Recognition.
Figure \ref{fig:flowChart} shows the block diagram of the system; the first
process is collecting stroke points then computing the curvature and possible
dominant points.  The following section provides a detailed description of each
step. The main block of the system remains the same but there are some
modifications in the segmentation algorithm to improve its efficiency. This study compares between using Traditional PSO and an Enhanced Particle swarm algorithm. Other modifications are explained in the details in
the next sections.
%  \begin{figure}
% 	\centering
% 		\includegraphics[width=3.5in,height=6in]{FlowChart.jpg}
% 	\caption{System flow chart} The chart shows the flow of process and data processed thought the system.
% 	\label{fig:flowChart}
% \end{figure}

@@

\subsection{Preprocessing}
\label{Prepross}

 Speed and time difference information were widely used in sketch recognition
systems \cite{earlyprocess}. Figure \ref{fig:orignalStroke} shows an example of
an input stroke drawn by users.  Figure \ref{fig:speed2Distance} shows the graph
of time difference and the speed for the stroke in Figure
\ref{fig:orignalStroke}. It was noted that time difference provides more
distinct maxima than the speed information. Agar et al. \cite{polygonfeedback31}
mentions that the pointing device (e.g., mouse) sampling rate is the reason for
this phenomenon. The points are sampled at regular time intervals as long as the
pointing device is moving. Therefore, there are no new detected points when the
pointing device is stationary. This leads to a nearly constant time difference
between samples while the pen is moving and large difference while the pen is
stationary (Figure \ref{fig:timediff}). Contrary to time difference information,
the speed information has a lot of noise in the data because usually  users draw
with variant speeds (Figure \ref{fig:speed2Distance}).  %Similarly, direction
%information is used as it provides better distinctive maxima for the corners
%than the curvature information \cite{meanshift10}.% Figure \ref{fig:curvatures}
%shows the direction and curvature graphs for the stroke drawn in Figure
\ref{fig:orignalStroke}

@@

 %Speed and time difference information were widely used in sketch understanding systems \cite{earlyprocess}. It was noted that time difference provides more distinct maxima than the speed information. Agar \cite{polygonfeedback31} mentions that the pointing device (i.e. the mouse) sampling rate is the reason for this phenomenon.
% % In the presented system, we computed time difference, direction, speed and
% curvature of each point along a stroke. The speed is calculated as $v=\Delta
% s/\Delta t$ where $t$ is the time difference between two points and $s$ is the
% length between them. The direction $d$ is calculated as the angle between two
% vectors and curvature is considered as the change in direction with respect to
% length i.e. $c= \Delta d/\Delta s$. Dominant points are characterized by lower
% speed values and higher curvature and direction values.

% After the system computes all the speed, time difference and curvature
% information it proceeds to detect the points with low velocity and high
% curvatures. Using simple differentiation to detect local extreme points resulted
% in false points due to the non smooth curves. Hence, the system adopted a
% process presented by \cite{earlyprocess}, where the mean of the curve is
% calculated. Then a threshold \textit{th} is used to separate the curve into
% regions $Region_i$; each region $Region_i$ is defined as a range of points,
% where the curve values are either above or below the threshold \textit{th}.
% Those regions are further processed to find the maximum point $Max(Region_i)$ of
% each region $Region_i$. The stroke points $p_i(x,y)$ that correspond to those
% maximum values are labeled as \textit{possible dominant points} $P_{pd}$. The
% system repeats this process for curvature, time difference and speed
% information. All the points labeled as possible dominant points $P_{pd}$ are
% saved into a single array. Figure \ref{fig:LabelsPPD} shows the particles
% labeled as Possible dominant point $P_{pd}$ by the preprocessing, it is noticed
% the redundancy of some $P_{pd}$ points. % (as shown are redundant)

@@

 \begin{figure}
	\centering
		\includegraphics[scale=0.48]{blockSmall.jpg}
	\caption{Segmentation Block}
	\label{fig:segblock}
\end{figure}
\subsection{Ellipse fitting}

This process tries to fit the stroke points into an ellipse arc; it starts with
computing the center of the stroke bounding box. The bounding box center point
is used as the first estimation of the center of the ellipse. The axes of the
ellipse are estimated as the $width/2$ and $height/2$ of the stroke bounding box
where $width$ is width and $height$ is height of the bonding box. The least
square fitting algorithm \cite{chernov} is used to minimize the fitting error of
the ellipse Equation (\ref{eq:circleFit})

\begin{equation}
E = \sum\limits_{i = 0}^N {\frac{{(x_i - x_0 )}}{{a^2 }}^2  + \frac{{(y_i - y_0
)}}{{b^2 }}^2  - 1}
\label{eq:circleFit}
\end{equation}

 where $N$ is number of points in the stroke, $a,b$ are the length of ellipse
axes, $x_0$ \& $y_0$ are the coordinates of the center point, $x_i$ \& $y_i$ are
the coordinates of point $i$ in the stroke. A list of new values for $x_0$ ,
$y_0$ ,$a$ and $b$ are generated randomly from the older values with small
increments after each loop.  After few iteration, the final fit error of the
estimated ellipse is reported. A modification to the system in \cite{mypaper} is
trying to add another measure to compute the efficiency of the final estimated
ellipse. Equation \ref{eq:circleError} ensures that the drawn percentage of
ellipse is considered. This eliminates fitting a line into very large ellipse
but leaves small ellipses to be fitted as a partial or full ellipse.

@@

Therefore, the system represents each particle $P_i$ by $P_i = \left\{ {p_{ij}
\left| {j = 1,2 \cdots N} \right.} \right\}$ where $p_{ij}$ has only two values
a)1 ($p_{ij}=1$); means that this point ($j$) is a dominate point, or b)
0($p_{ij}=0$) means this point ($j$) is not a dominate point.
Figure\ref{fig:pso1} shows a particle encoding in the DPSO system.

	\item[Fitness function:] The fitness function and error calculation are
different in each of the two \textit{DPSO} algorithms.
	\end{description}
\subsubsection{\textit{Polygon approximation\textsl{AlgS1}}}
%\\
The approximation error is computed by the equation \ref{eq:ErrorSwarm1}. The
graphical meaning of the error is shown in Figure\ref{fig:pso1}.%\ref{fig:LabelsPSO}.
\begin{equation}
E=\sum\nolimits_{i = 0}^M e ( \widehat{x_ix_{i+1}},\overline{x_i x_{i+1}})
\label{eq:ErrorSwarm1}
\end{equation} where the arc $\widehat{x_ix_j}$ is defined as the consecutive
set of points from point $x_i$ to point $x_{j}$ as in $x_i,x_{i+1} \cdots,x_j$.
The line $\overline{x_i x_j} $ as the straight line connecting point $x_i$ to
point $x_j$, $M$ is the number of dominant points in this solution as generated
by the swarm algorithm. The error $e ( \widehat{x_ix_j},\overline{x_i x_j})$ is
computed as the sum of squared perpendicular distance from every point along the
arc $\widehat{x_ix_j}$ to the line $\overline{x_i x_j}$. The fitness is computed
using equation %\ref{eq:fitnessSwarm1}
\begin{equation}
\max fitness(p_i ) = \left\{ {\begin{array}{*{20}c}
   { - E/\varepsilon N} & {ifE > \varepsilon ,}  \\
   {D/\sum\limits_{j = 1}^N {p_{ij} } } & {otherwise}  \\
\end{array}} \right.
\label{eq:fitnessSwarm1}
\end{equation} where $N$ is the number of points in the stroke, $D$ is the
number of points in the solution that was previously labeled as a possible
dominant point ($P_{pd}$), $E$ is the computed error and $\varepsilon$ is the
error threshold. It should be noticed that when the error is larger than the
threshold $\varepsilon$ the fitness is given a -ve value to lower the fitness
value of the solution. Otherwise the system favors the lower number of vertices.

@@

\subsection{Recognition}
\label{sec:Recognition}
After the user draws all strokes of the symbol, the set of un-recognized strokes
is grouped together along with their segmentation as input to the feature
extraction process. A composite set of features is used to generate a single
feature vector. The following list describes each feature in details.
%\begin{description}
\subsubsection{Structural and geometrical Features (FS1)}
  They are features that define the structure of the geometrical symbol such as;
 \begin{itemize}
	 \item \emph{Segments:} The number of segments in the symbol.
	 \item \emph{Strokes:} The number of strokes or partial strokes that
created the symbols.
		\item  \emph{Primitives:} The number of primitives in the
symbol. This feature helps when identifying  symbols with mixed geometric
primitives such as cylinders and callouts.
		\item \emph{Curves:} The number of curves or ellipses in the
symbol.
		\item \emph{Lines:} The number of lines in the symbol.
		\item \emph{Perpendicular}\emph{lines:} The number of
perpendicular lines.
		\item \emph{Parallel}\emph{lines:} The number of parallel lines.
		\item \emph{Intersections:} The number of intersection between
lines and curves.
		\item \emph{T intersections:} The number of T intersections.
		\item \emph{L intersections:} The number of L intersections.
		\item \emph{X intersections:} The number of X intersections.

@@

\subsubsection{Statistical Features (FS3)}
These features use moments and statistical properties such as;
\begin{itemize}
	\item \emph{Zernike moments:} The zernike moments are orthogonal moments
which define the distribution of point in the input space. They are invariant to
both reflection and rotation of the input shape \cite{ZerMomentOrthogonal}. %The
zernike moment were used in \cite{HeloiseBeautification} t.
\end{itemize}
\subsubsection{Composite Features (FS4)}
 These features are composites of different geometrical and statistical
properties of the symbol.
	\begin{itemize}
\item \emph{Size Ratio:} The ratio between width and height of the symbol.
	\item \emph{Ink density:} The density of points inside its bounding
box\cite{GeometryAndDomain102}.
 	\item \emph{Convex Hull Area:} The area of convex hull with respect to
area of bounding box of symbol.
	\item \emph{Convex Hull Perimeter:} The perimeter of convex hull with
respect to total length of symbol.
		\item \emph{Mean Centroidal radius:} The Mean of the centroidal
radius is the distance from each point in the symbol to the center of gravity.
	\item \emph{Mean Time difference:} The mean of the time difference
between each two successive points in the symbol.
  \end{itemize}

 After computing the features, the symbol is introduced to the classifier as a
feature vector. The system uses Support Vector Machine (SVM) classifier with
Linear kernel \cite{libsvm}. An OVO classifier structure (object versus object)
is used to handle multi-class problem.% A validation algorithm is used to choose
the Kernel variables parameters $c, \gamma$.

@@

 \begin{figure}
	\centering
	\subfigure[Algorithm comparison on HS-DS]{\label{fig:test1} \includegraphics[scale=0.4]{images/testAlg.jpg}}
 		\subfigure[Algorithm comparison on LD-DS]{ 	 	\label{fig:testLD} \includegraphics[scale=0.4]{images/LDAlg.jpg}}
 		\subfigure[Algorithm comparison on EL-DS]{	 	\label{fig:testEL}  \includegraphics[scale=0.4]{images/AlgEL.jpg}}
	 	\caption{ The recognition rate of different algorithms. }
 \end{figure}
\subsection{Swarm Algorithms Experiments}
Another set of tests were performed on the system to enhance non-ellipse fitting
algorithms. The effect of the number of particles, the error thresholds and
various other parameters were investigated. The results of these tests are shown
in Figures \ref{fig:swarmtesting} and   \ref{fig:swarmtesting2}. Figure
\ref{fig:swarmtesting} shows that as the number of swarm iterations increase the
error value decrease (error of the segmentation algorithm). The final number of
particles and maximum swarm iterations used in the system was based on a
tradeoff between error achieved versus the computation time. The effect of other
algorithm parameters such as $c_1$, $\omega$ (mentioned in
Section\ref{sec:ParticleSwarmAlgorithm}) is tested using similar tests (Figure
\ref{fig:swarmtesting2},\ref{fig:swarmwtest}).%, $c_2$

>>>>

\section{Enhanced Particle Swarm}
\label{sec:EnhancedSwarmAlgorithm}

In the above basic PSO, the velocity update is based on three terms an inertia term, local cognitive ($l_{best}$) term and global cognitive ($g_{best}$)term. The particle local cognitive term is associated with the particle own history without consideration of other swarm particles locations and histories which leads to a lake of diversity and sub optimum solutions. Therefore, we are trying to add more social influence in the particle movement by using the weighted difference of the position vectors of any other two distinct particles randomly chosen from the swarm to modify the local cognitive term. The modified term adds social behavior of other particle and keeps diversity while searching for the global optimum. The new social cognitive term of the particle is shifted to the new location only if it yields a fitness value better than obtained using the standard PSO\cite{stockPaper}.

@@

\section{System Discribtion  }\label{Sysdisc}
%\label{sec:application}
%The modified swarm algorithm is used to improve the performance of the sketch recognition system in \cite{mypaper}.
 The sketch recognitions system is divided into three main steps 1) preprocessing, 2) segmentation and 3) Recognition. Figure \ref{fig:flowChart} shows the block diagram of the system; the first process is collecting stroke points then computing the curvature and possible dominant points.  The following section provides a detailed description of each step. The main block of the system remains the same but there are some modifications in the segmentation algorithm to improve its efficiency. This study compares between using Traditional PSO and an Enhanced Particle swarm algorithm. Other modifications are explained in the details in the next sections.
%  \begin{figure}
% 	\centering
% 		\includegraphics[width=3.5in,height=6in]{FlowChart.jpg}
% 	\caption{System flow chart} The chart shows the flow of process and data processed thought the system.
% 	\label{fig:flowChart}
% \end{figure}

@@

\subsection{Preprocessing}
\label{Prepross}
  Speed and time difference information were widely used in sketch recognition systems \cite{earlyprocess}. Figure \ref{fig:orignalStroke} shows an example of an input stroke drawn by users.  Figure \ref{fig:speed2Distance} shows the graph of time difference and the speed for the stroke in Figure \ref{fig:orignalStroke}. It was noted that time difference provides more distinct maxima than the speed information. Agar et al. \cite{polygonfeedback31} mentions that the pointing device (e.g., mouse) sampling rate is the reason for this phenomenon. The points are sampled at regular time intervals as long as the pointing device is moving. Therefore, there are no new detected points when the pointing device is stationary. This leads to a nearly constant time difference between samples while the pen is moving and large difference while the pen is stationary (Figure \ref{fig:timediff}). Contrary to time difference information, the speed information has a lot of noise in the data because usually  users draw with variant speeds (Figure \ref{fig:speed2Distance}).  %Similarly, direction
%information is used as it provides better distinctive maxima for the corners
%than the curvature information \cite{meanshift10}.% Figure \ref{fig:curvatures}
%shows the direction and curvature graphs for the stroke drawn in Figure
\ref{fig:orignalStroke}

@@

@@

 \begin{figure}
	\centering
		\includegraphics[scale=0.48]{blockSmall.jpg}
	\caption{Segmentation Block}
	\label{fig:segblock}
\end{figure}
\subsection{Ellipse fitting}  This process tries to fit the stroke points into an ellipse arc; it starts with computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the center of the ellipse. The axes of the ellipse are estimated as the $width/2$ and $height/2$ of the stroke bounding box where $width$ is width and $height$ is height of the bonding box. The least square fitting algorithm \cite{chernov} is used to minimize the fitting error of the ellipse Equation (\ref{eq:circleFit})

\begin{equation}
E = \sum\limits_{i = 0}^N {\frac{{(x_i - x_0 )}}{{a^2 }}^2  + \frac{{(y_i - y_0
)}}{{b^2 }}^2  - 1}
\label{eq:circleFit}
\end{equation}
  where $N$ is number of points in the stroke, $a,b$ are the length of ellipse axes, $x_0$ \& $y_0$ are the coordinates of the center point, $x_i$ \& $y_i$ are the coordinates of point $i$ in the stroke. A list of new values for $x_0$ , $y_0$ ,$a$ and $b$ are generated randomly from the older values with small increments after each loop.  After few iteration, the final fit error of the estimated ellipse is reported. A modification to the system in \cite{mypaper} is trying to add another measure to compute the efficiency of the final estimated ellipse. Equation \ref{eq:circleError} ensures that the drawn percentage of ellipse is considered. This eliminates fitting a line into very large ellipse but leaves small ellipses to be fitted as a partial or full ellipse.

@@

Therefore, the system represents each particle $P_i$ by $P_i = \left\{ {p_{ij}
\left| {j = 1,2 \cdots N} \right.} \right\}$ where $p_{ij}$ has only two values
a)1 ($p_{ij}=1$); means that this point ($j$) is a dominate point, or b)
0($p_{ij}=0$) means this point ($j$) is not a dominate point.
Figure\ref{fig:pso1} shows a particle encoding in the DPSO system.
	\item[Fitness function:] The fitness function and error calculation are
different in each of the two \textit{DPSO} algorithms.
	\end{description}
\subsubsection{\textit{Polygon approximation\textsl{AlgS1}}}
%\\
The approximation error is computed by the equation \ref{eq:ErrorSwarm1}. The
graphical meaning of the error is shown in Figure\ref{fig:pso1}.%\ref{fig:LabelsPSO}.
\begin{equation}
E=\sum\nolimits_{i = 0}^M e ( \widehat{x_ix_{i+1}},\overline{x_i x_{i+1}}) \label{eq:ErrorSwarm1} \end{equation} 
where the arc $\widehat{x_ix_j}$ is defined as the consecutive set of points from point $x_i$ to point $x_{j}$ as in $x_i,x_{i+1} \cdots,x_j$. The line $\overline{x_i x_j} $ as the straight line connecting point $x_i$ to point $x_j$, $M$ is the number of dominant points in this solution as generated by the swarm algorithm. The error $e ( \widehat{x_ix_j},\overline{x_i x_j})$ is computed as the sum of squared perpendicular distance from every point along the arc $\widehat{x_ix_j}$ to the line $\overline{x_i x_j}$. The fitness is computed
using equation %\ref{eq:fitnessSwarm1}
\begin{equation}
\max fitness(p_i ) = \left\{ {\begin{array}{*{20}c}
   { - E/\varepsilon N} & {ifE > \varepsilon ,}  \\
   {D/\sum\limits_{j = 1}^N {p_{ij} } } & {otherwise}  \\ 
   \end{array}} \right. \label{eq:fitnessSwarm1} \end{equation}

   where $N$ is the number of points in the stroke, $D$ is the number of points in the solution that was previously labeled as a possible dominant point ($P_{pd}$), $E$ is the computed error and $\varepsilon$ is the error threshold. It should be noticed that when the error is larger than the threshold $\varepsilon$ the fitness is given a -ve value to lower the fitness value of the solution. Otherwise the system favors the lower number of vertices.

@@

\subsection{Recognition}
\label{sec:Recognition}

After the user draws all strokes of the symbol, the set of un-recognized strokes is grouped together along with their segmentation as input to the feature extraction process. A composite set of features is used to generate a single feature vector. The following list describes each feature in details.
%\begin{description}

\subsubsection{Structural and geometrical Features (FS1)}
  They are features that define the structure of the geometrical symbol such as;
 \begin{itemize}
	 \item \emph{Segments:} The number of segments in the symbol.
	 \item \emph{Strokes:} The number of strokes or partial strokes that
created the symbols.
		\item  \emph{Primitives:} The number of primitives in the
symbol. This feature helps when identifying  symbols with mixed geometric
primitives such as cylinders and callouts.
		\item \emph{Curves:} The number of curves or ellipses in the
symbol.
		\item \emph{Lines:} The number of lines in the symbol.
		\item \emph{Perpendicular}\emph{lines:} The number of
perpendicular lines.
		\item \emph{Parallel}\emph{lines:} The number of parallel lines.
		\item \emph{Intersections:} The number of intersection between
lines and curves.
		\item \emph{T intersections:} The number of T intersections.
		\item \emph{L intersections:} The number of L intersections.
		\item \emph{X intersections:} The number of X intersections.

@@

\subsubsection{Statistical Features (FS3)}
These features use moments and statistical properties such as;
\begin{itemize}
 \item \emph{Zernike moments:} The zernike moments are orthogonal moments which define the distribution of point in the input space. They are invariant to both reflection and rotation of the input shape \cite{ZerMomentOrthogonal}. %The zernike moment were used in \cite{HeloiseBeautification} t.
\end{itemize}
\subsubsection{Composite Features (FS4)}
 These features are composites of different geometrical and statistical
properties of the symbol.
	\begin{itemize}
\item \emph{Size Ratio:} The ratio between width and height of the symbol.
	\item \emph{Ink density:} The density of points inside its bounding
box\cite{GeometryAndDomain102}.
 	\item \emph{Convex Hull Area:} The area of convex hull with respect to
area of bounding box of symbol.
	\item \emph{Convex Hull Perimeter:} The perimeter of convex hull with
respect to total length of symbol.
		\item \emph{Mean Centroidal radius:} The Mean of the centroidal
radius is the distance from each point in the symbol to the center of gravity.
	\item \emph{Mean Time difference:} The mean of the time difference
between each two successive points in the symbol.
  \end{itemize} 
  After computing the features, the symbol is introduced to the classifier as a feature vector. The system uses Support Vector Machine (SVM) classifier with Linear kernel \cite{libsvm}. An OVO classifier structure (object versus object) is used to handle multi-class problem.% A validation algorithm is used to choose the Kernel variables parameters $c, \gamma$.

@@

 \begin{figure}
	\centering
	\subfigure[Algorithm comparison on HS-DS]{\label{fig:test1} \includegraphics[scale=0.4]{images/testAlg.jpg}}
 		\subfigure[Algorithm comparison on LD-DS]{ 	 	\label{fig:testLD} \includegraphics[scale=0.4]{images/LDAlg.jpg}}
 		\subfigure[Algorithm comparison on EL-DS]{	 	\label{fig:testEL}  \includegraphics[scale=0.4]{images/AlgEL.jpg}}
	 	\caption{ The recognition rate of different algorithms. }
 \end{figure} 
 \subsection{Swarm Algorithms Experiments}

 Another set of tests were performed on the system to enhance non-ellipse fitting algorithms. The effect of the number of particles, the error thresholds and various other parameters were investigated. The results of these tests are shown in Figures \ref{fig:swarmtesting} and   \ref{fig:swarmtesting2}. Figure \ref{fig:swarmtesting} shows that as the number of swarm iterations increase the error value decrease (error of the segmentation algorithm). The final number of particles and maximum swarm iterations used in the system was based on a tradeoff between error achieved versus the computation time. The effect of other algorithm parameters such as $c_1$, $\omega$ (mentioned in Section\ref{sec:ParticleSwarmAlgorithm}) is tested using similar tests (Figure \ref{fig:swarmtesting2},\ref{fig:swarmwtest}).%, $c_2$
